---
title: "R and the command line, sweave, extras..."
author: "Caitlin Simopoulos"
date: '2017-05-19'
output: html_document
highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Command line

While R Studio is great for learning R, sometimes you just want to write some code and let it run.
I like to call this "time to leave early", or "time for an extended coffee break".
You can also write code in any text editor you want (VI, R Studio, Atom, SublimeTest...) and run your code similar to how you would run python/perl.
To run your code you can use a command like:

```{r, engine = 'bash', eval = FALSE}
R --vanilla < yourRScript.R > yourOutfile.out

# --vanilla starts R fresh
# yourOutfile.out is everything that would normally print out on the console, good for debugging
```

## Some key functions...

Because you're not interactively using R, you will need to be printing, writing CSVs, and saving R objects often. 
I think one of my favourite things to do, especially when working with large files that take a long time to do anything to, is using R's save() function.
You can either save a single, or subset of objects using save(), or can save your entire workspace using save.image(). 
Conventionally, we save these images as ".RData" files.
```{r, cache=TRUE, echo=FALSE}
setwd("~/Dropbox/R_teaching/R_basics/")
```

```{r, cache=TRUE}
# let's say you wanted to run something overnight, and then wanted to access the results IN R, without having to save it as a CSV.

data(iris)
pca <- prcomp(iris[,1:4])
save(pca, file="iris_pca.RData")
save.image(file="everything.RData")
load("everything.RData") # literally everything that's in your workspace will be saved...you don't always need this!

```

# More advanced useful functions

```{r, eval=FALSE}

which()
# because who needs if else statements...
ifelse()
# for loops are so passe
apply()
lapply()
```

